# Checking assumptions and data transformations {#assumptions_trans}

```{r echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")
```

**Tutorial learning objectives**

* Learn how to check the normality assumption
  + Normal quantile plots
  + Shapiro-Wilk test for normality  
    
* Learn how to check the equal variance assumption
  + Levene's Test 
    
* Learn how to transform the response variable to help meet assumptions   
  + log-transform
  + Dealing with zeroes  
  + log bases
  + back-transforming log data
  + logit transform
  + back-transforming logit data
  + when to back-transform?

Most statistical tests, such as the <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Chi-square-($\chi$^2^)-goodness-of-fit-test">$\chi$^2^ goodness of fit test</a>, <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Chi-square-($\chi$^2^)-contingency-test">the $\chi$^2^ contingency test</a>, *t*-test, <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Analysis-of-variance">ANOVA</a>, Pearson correlation, and least-squares <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Linear-regression">regression</a>, have <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Assumptions">assumptions</a> that must be met. For example, the one-sample *t*-test requires that the variable is normally distributed in the population, and least-squares regression requires that the residuals from the regression be normally distributed. In this tutorial we'll learn ways to check the assumption that the variable is normally distributed in the population.   

We'll also learn how transforming a variable can sometimes help satisfy assumptions, in which case the analysis is conducted on the transformed variable. 

## Load packages and import data {#assumptions_packages_data}

Load the usual packages, and `broom`, which has been used in some tutorials:

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(skimr)
library(broom)
library(biol202)
```

And we need these two packages also: `car`, `boot`.  Install these if you don't have them (as per instructions in a previous [tutorial](01-getting-started.qmd#package_install)), then load them:

```{r, message = FALSE, warning = FALSE}
library(car)  
library(boot)
```

The "marine.csv" dataset is discussed in example 13.1 in the text book.  The "flowers.csv" dataset is described below. The "students.csv" data include data about BIOL202 students from a few years ago.  

Let's make sure to treat any categorical variables as factor variables, using the "stringsAsFactors = T" argument:

```{r}
data(marine)
data(flowers)
data(students)
```

Explore the `marine` and `flowers` datasets:  

```{r skim_marine}
marine %>%
  skim_without_charts()
```

```{r skim_flowers}
flowers %>%
  skim_without_charts()
```

## Checking the normality assumption {#checknorm_assum}

Statistical tests such as the one-sample *t* test assume that the response variable of interest is normally distributed in the population.  

Many biological variables are known to be normally distributed in the population, but for some variables we can't be sure.  Given a proper random sample from the population, of sufficient sample size, we can assume that the frequency distribution of our sample data will, to reasonable degree, reflect the frequency distribution of the variable in the population.

Importantly, tests such as the one-sample *t* test are somewhat **robust** to minor violations of this assumption.  Nevertheless, it is best practice to be transparent in testing the assumption, i.e. showing how it was tested and exactly what was found.

### Normal quantile plots {#quantile_plots}

The most straightforward way to check the normality assumption is to visualize the data using a **normal quantile plot**.  

The `ggplot2` package has plotting functions for this, called `stat_qq` and `stat_qq_line`:

```
?stat_qq
?stat_qq_line
```

For details about what Normal Quantile Plots are, and how they're constructed, consult this informative [link](https://data.library.virginia.edu/understanding-q-q-plots/). 

::: {.callout-note}
If the frequency distribution were normally distributed, points would fall close to the straight line in the normal quantile plot.
:::

Check out this example showing simulated data drawn from a normal distribution:

```{r echo = F}
set.seed(1523)
temp.norm <- data.frame(tempdata = rnorm(32))
```

```{r echo = F, fig.width = 5, fig.height = 4, fig.cap = "Example of a normal quantile plot for a variable that is normally distributed."}

  ggplot(data = temp.norm, aes(sample = tempdata)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  xlab("Normal quantile") +
  theme_bw()
```

Now we'll use the `marine` dataset and its variable called `biomass_ratio` to illustrate.

We'll first construct a histogram as you've [learned previously](05-visualize-one-variable.qmd#vis_histogram), just to see how the shape of the frequency distribution relates to the pattern seen in the normal quantile plot.

```{r echo = TRUE, fig.cap = "The frequency distribution of the 'biomass ratio' of 32 marine reserves.", fig.width = 4, fig.height = 3}

marine %>%
  ggplot(aes(x = biomass_ratio)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightgrey", 
                 boundary = 0, closed = "left") +
  xlab("Biomass ratio") +
  ylab("Frequency") +
  theme_bw()
```

Notice that the distribution is quite right-skewed (or "postively skewed"). 

Now the quantile plot:

```{r fig.cap = "Normal quantile plot of the 'biomass ratio' of 32 marine reserves.", fig.width = 4, fig.height = 3}
marine %>% 
 ggplot(aes(sample = biomass_ratio)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  ylab("Biomass ratio") +
  xlab("Normal quantile") +
  theme_bw()
```

Notice that in the "aes" argument we use "sample = biomass_ratio".  This is new, and is only required for the normal quantile plot, specifically the subsequent `stat_qq` and `stat_qq_line` functions.

Notice that normal quantile plot shows points deviating substantially from the straight line in the top-right part of the plot, and this corresponds to the right-skew in the histogram. 

Clearly, the frequency distribution of the `biomass_ratio` variable does not conform to a normal distribution.  

Here's an example statement one could make when checking this assumption:

>The assumption of normality was checked visually using a normal quantile plot, which showed that the data were clearly not normally distributed.

::: {.callout-note}
**Important: **
Egregious deviations from the normality assumption will be clearly evident in normal quantile plots (as in the example above). If it is difficult to tell whether the data are normally distributed, then they probably are OK (at least sufficiently with respect to the assumption).  
:::

### Shapiro-Wilk test for normality {#shapirottest}

Although graphical assessments are usually sufficient for checking the normality assumption, one can conduct a formal statistical test of the <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Null-hypothesis">null hypothesis</a> that the data are sampled from a population having a normal distribution. The test is called the **Shapiro-Wilk test**.  

The Shapiro-Wilk test is a type of goodness-of-fit test.  

::: {.callout-note}
Sometimes the **Shapiro-Wilk test** is applied in a <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Hypothesis-testing">hypothesis testing</a> framework, but when it is applied as part of checking assumptions for another statistical test (like we're doing here), one does not need to present it in a hypothesis test framework. However, the implied null hypothesis is that "The data are sampled from a population having a normal distribution", and one does interpret the resulting *P*-value in the same way as usual, i.e. in relation to an $\alpha$ level (see below).
:::

We'll make use of the `shapiro.test` function from base R stats package:

```
shapiro.test
```

Let's use this on the `biomass_ratio` variable in the "marine" dataset. We'll assign the output to an object called "shapiro.result", then we'll have a look at the results.

::: {.callout-note}
The output from the `shapiro.test` function is not "tidy", so we will use the `tidy` function from the `broom` package to make it tidy.
:::

Here we go: we provide the function with the tibble name ("marine") and the variable of interest after a "$":
```{r normtest}
shapiro.result <- shapiro.test(marine$biomass_ratio)
```

Now tidy the output:

```{r tidynormal}
shapiro.result.tidy <- tidy(shapiro.result)
```

Now look at the results:

```{r normtest_results}
shapiro.result.tidy
```

The tidy object includes:  

* The value of the test statistic for the Shapriro-Wilk test (although it is not shown, this test statistic is indicated with a "W")
* The *P*-value associated with the test ("p.value")
* The name of the test used ("method")

Given that the *P*-value is less than a conventional $\alpha$ level of 0.05, the test is telling us that the data do not conform to a normal distribution.  Of course we already knew that from our visual assessments!

Here's an example statement:

>The assumption of normality was checked visually using a normal quantile plot, and a Shapiro-Wilk test, which revealed evidence of non-normality (Shapiro-Wilk test, W = `r round(shapiro.result.tidy$statistic, 2)`, *P*-value < 0.001).

::: {.callout-note}
**Important: **
Visual assessments of normality are preferred, because the outcome of the Shapiro-Wilk test is sensitive to sample size: a small sample size will often yield a false negative, whereas very large sample sizes could yield false positives more than it should.
:::

## Checking the equal-variance assumption {#levenetest}

Some statistical tests, such as the 2-sample *t*-test and ANOVA, assume that the <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Variance">variance ($\sigma$)</a> of the numeric response variable is the same among the populations being compared.

For this we use the **Levene's test**, which we implement using the `leveneTest` function from the `car` package:

```
?leveneTest
```

Like the Shapiro-Wilk test, the Levene's test can be applied in a hypothesis testing framework, but when it is used to evaluate the equal-variance assumption, it need not be.

Nevertheless, the _implied_ null hypothesis is that the variance of the numeric response variable is the same among the populations being compared. So in the case where a numeric variable is being compared among two groups, then the implied null hypothesis is that ($\sigma$(1) = $\sigma$(2). The usual $\alpha$ level of 0.05 can be used.

We'll use the "students" dataset, and check whether `height_cm` exhibits equal variance among students with different dominant eyes (left or right).  

Let's look at the code, and explain after:

```{r do_levene}
height.vartest <- leveneTest(height_cm ~ dominant_eye, data = students)
```

::: {.callout-note}
If you get a warning about a variable being "coerced to factor", that's OK! It is simply telling you that it took the categorical variable and treated it as a 'factor' variable.
:::

In the code chunk above we: 

* assign the results to a new object called "height.vartest"
* use the `leveneTest` function, in which the arguments are:
  + the numeric response variable ("height_cm")
  + then the "~" symbol
  + then the categorical variable "dominant_eye"
  + then the "data = students" specifies the data object name

::: {.callout-note}
**TIP: **
The argument to the `leveneTest` function that is in the form $Y$ ~ $X$ is one we'll use several times.
:::

Let's look at the results:

```{r levene_results}
height.vartest
```

The results include: 

* the degrees of freedom for the test
* the value of the test statistic "F"
* the *P*-value associated with the test statistic  

For the student height example, the *P*-value is greater than the standard $\alpha$ of 0.05, so there's no evidence against the assumption of equal variance.

A reasonable statement would be:

>"A Levene's test showed no evidence against the assumption of equal variance (*F* = `r round(unlist(height.vartest)[3], 2)`; *P*-value = `r round(unlist(height.vartest)[5], 3)`)."

::: {.callout-note}
**TIP: **
As part of your main statistical test, such as a two-sample *t*-test or an ANOVA, you would typically provide a graph such as a [stripchart or violin plot](07-visualize-two-variables.qmd#numeric_vs_cat) to visualize how the numeric response variable varies among the groups of the categorical explanatory variable.  Such plots may often reveal that the spread of values of the response variable varies considerably among the groups, underscoring the need to check the equal variance assumption.  
:::

## Data transformations {#transform}

Here we learn how to <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Data-transformation">transform</a> numeric variables using two common methods:  

* log-transform
* logit-transform  

There are many other types of transformations that can be performed, some of which are described in **Chapter 13** of the course text book.  

::: {.callout-note}
Contrary to what is suggested in the text, it is better to use the "logit" transformation rather than the "arcsin square-root" transformation for <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Proportion">proportion</a> or percentage data, as described in this [article](http://onlinelibrary.wiley.com/doi/10.1890/10-0340.1/abstract) by Warton and Hui (2011).
:::

### Log-transform {#log_transform}

When one observes a right-skewed frequency distribution, as seen here in the marine biomass ratio data, a log-transformation often helps.  

```{r echo = TRUE, fig.cap = "The frequency distribution of the 'biomass ratio' of 32 marine reserves.", fig.width = 4, fig.height = 3}

marine %>%
  ggplot(aes(x = biomass_ratio)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "lightgrey", 
                 boundary = 0, closed = "left") +
  xlab("Biomass ratio") +
  ylab("Frequency") +
  theme_bw()
```

To log-transform the data, simply create a new variable in the dataset using the `mutate` function (from the `dplyr` package) that we've seen before.  Here we'll call our new variable `logbiomass`, and use the `log` function to take the **natural log** of the "biomass_ratio" variable.

We'll assign the output to the same, original "tibble" called "marine":

```{r logtrans_marine3}
marine <- marine %>%
  mutate(logbiomass = log(biomass_ratio))
```

Alternatively, you could use this (less tidy) code to get the same result:

```
marine$logbiomass <- log(marine$biomass_ratio)
```

::: {.callout-note}
If your variable includes zeros, then you'll need to take extra steps, as described in the next [section](#zeroes).
:::

Let's look at the tibble now:

```{r logtrans_marine_see}
marine
```

Now let's look at the histogram of the log-transformed data:

```{r echo = TRUE, fig.cap = "The frequency distribution of the 'biomass ratio' of 32 marine reserves (log-transformed).", fig.width = 5, fig.height = 4}

marine %>%
  ggplot(aes(x = logbiomass)) +
  geom_histogram(binwidth = 0.25, color = "black", fill = "lightgrey", 
                 boundary = -0.25) +
  xlab("Biomass ratio (log-transformed)") +
  ylab("Frequency") +
  theme_bw()
```

Now the quantile plot:

```{r fig.cap = "Figure: Normal quantile plot of the 'biomass ratio' of 32 marine reserves (log-transformed).", fig.width = 4, fig.height = 3}

marine %>%
  ggplot(aes(sample = logbiomass)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  ylab("Biomass ratio (log)") +
  xlab("Normal quantile") +
  theme_bw()
```

The log-transform definitely helped, but the distribution still looks a bit wonky: several of the points are quite far from the line.  

Just to be sure, let's conduct a Shapiro-Wilk test, using an <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Significance-level-($\alpha$)">$\alpha$ level</a> of 0.05, and remembering to tidy the output:

```{r normtest22}
shapiro.log.result <- shapiro.test(marine$logbiomass)
shapiro.log.result.tidy <- tidy(shapiro.log.result)
shapiro.log.result.tidy
```

The *P*-value is greater than 0.05, so we'd conclude that there's no evidence against the assumption  that these data come from a normal distribution.  

For this example a reasonable statement would be:

>Based on the normal quantile plot (Fig. 18.9), and a Shapiro-Wilk test, we found no evidence against the normality assumption (Shapiro-Wilk test, W = `r round(shapiro.log.result.tidy$statistic, 2)`, *P*-value = `r round(shapiro.log.result.tidy$p.value, 3)`).

You could now proceed with the statistical test (e.g. one-sample *t*-test) using the **transformed** variable. 

### Dealing with zeroes {#zeroes}

If you try to log-transform a value of zero, R will return a `-Inf` value.  

In this case, you'll need to add a constant (value) to each observation, and convention is to simply add 1 to each value prior to log-transforming.  

In fact, you can add any constant that makes the data conform best to the assumptions once log-transformed. The key is that you must add the same constant to every value in the variable.   

You then conduct the analyses using these newly transformed data (which had 1 added prior to log-transform), remembering that after back-transformation (see below), you need to subtract 1 to get back to the original scale.  

**Example code showing how to check for zeroes**

We'll create a dataset to work with called "apples".

Don't worry about learning this code... 

```{r createdataset}
set.seed(345)
apples <- as_tibble(data.frame(biomass = rlnorm(n = 14, meanlog = 1, sdlog = 0.7)))
apples$biomass[4] <- 0
```

Here's the resulting dataset, which includes a variable "biomass" that would benefit from log-transform:

```{r}
apples
```

Here's a quantile plot of the biomass variable:

```{r echo = F, fig.width = 5, fig.height = 4, fig.cap = "Normal quantile plot of made-up biomass data."}

apples %>% 
  ggplot(aes(sample = biomass)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  xlab("Normal quantile") +
  theme_bw()
```

Let's first see what happens when we try to log-transform the "biomass" variable:

```{r}
log(apples$biomass)
```

Notice we get a "-Inf" value.

The following code tallies the number of observations in the "biomass" variable that equal zero. 

**If this sum is greater than zero**, then you'll need to add a constant to all observations when transforming.

```{r}
sum(apples$biomass == 0)
```

So we have one value that equals zero.

So let's add a 1 to each observation during the process of log-transforming:

```{r}
apples <- apples %>%
  mutate(logbiomass_plus1 = log(biomass + 1))
```

Notice that we name the new variable "logbiomass_plus1" in a way that indicates we've added 1 prior to log-transforming, and that in the `log` calculation we've used "biomass + 1".

Let's see the result:

```{r}
apples
```

Notice that we still have a zero in the newly created variable, **AFTER** having transformed, because for that value we calculated the log of "1" (which equals zero).  That's OK!

```{r echo = F, fig.width = 5, fig.height = 4, fig.cap = "Normal quantile plot of made-up biomass data, log-transformed."}

apples %>% 
  ggplot(aes(sample = logbiomass_plus1)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  xlab("Normal quantile") +
  theme_bw()
```

That's a bit better.  We would now use this new variable in our analyses (assuming it meets the normality assumption). 

### Log bases {#logbases}

The `log` function calculates the natural logarithm (base *e*), but related functions permit any base:  

```
?log
```

For instance, `log10` uses log base 10:

```{r logtrans_marine1}
marine <- marine %>%
  mutate(log10biomass = log10(biomass_ratio))
marine
```

Or the alternative code: 

```
marine$log10biomass <- log10(marine$biomass_ratio)
```

### Back-transforming log data {#back_transform}

In order to back-transform data that were transformed using the natural logarithm (`log`), you make use of the `exp` function:  

```
?exp
```

Let's try it, creating a new variable in the "marine" dataset so we can compare to the original "biomass_ratio" variable:

First, back-transform the data and store the results in a new variable within the data frame:  

```{r logtrans_marine2}
marine <- marine %>%
  mutate(back_biomass = exp(logbiomass))
```

Now have a look at the first few lines of the tibble (selecting the original "biomass_ratio" and new "back_biomass" variables) to see if the data values are identical, as they should be:  

```{r mrd}
marine %>%
  select(biomass_ratio, back_biomass)
```

Yup, it worked!   

If you had added a 1 to your variable prior to log-transforming, then the code would be:

```
marine <- marine %>%
  mutate(back_biomass = exp(logbiomass) - 1)
```

Notice the minus 1 comes after the `exp` function is executed.

If you had used the log base 10 transformation, then the code to back-transform is as follows:  

```{r log10_back}
10^(marine$log10biomass)
```

The `^` symbol stands for "exponent". So here we're calculating 10 to the exponent *x*, where *x* is each value in the dataset.  

### Logit transform {#logit_transform}

Variables whose data represent proportions or percentages are, by definition, not drawn from a normal distribution: they are bound by 0 and 1 (or 0 and 100%). They should therefore be **logit-transformed**.  

The `boot` package includes both the `logit` function and the `inv.logit` function, the latter for back-transforming.  

However, the `logit` function that is in the `car` package is better, because it accommodates the possibility that your dataset includes a zero and / or a one (equivalently, a zero or 100 percent), and has a mechanism to deal with this properly.  

The `logit` function in the `boot` package does not deal with this possibility for you.  

However, the `car` package does not have a function that will back-transform logit-transformed data.  

This is why we'll **use the `logit` function from the `car` package, and the `inv.logit` function from the `boot` package!**  

Let's see how it works with the `flowers` dataset, which includes a variable `prop_fertile` that describes the proportion of seeds produced by individual plants that were fertilized.  

Let's visualize the data with a normal quantile plot:

```{r fig.cap = "Normal quantile plot of the proportion of seeds fertilized on 30 plants (left) and the corresponding normal quantile plot (right)", fig.width = 4, fig.height = 3}

flowers %>%
  ggplot(aes(sample = prop_fertile)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  ylab("Proportion of seeds fertilized") +
  xlab("Normal quantile") +
  theme_bw()
```

Clearly not normal! 

Now let's logit-transform the data.  

To ensure that we're using the correct `logit` function, i.e. the one from the `car` package and NOT from the `boot` package, we can use the `::` syntax, with the package name preceding the double-colons, which tells R the correct package to use.

```{r brd}
flowers <- flowers %>%
  mutate(logitfertile = car::logit(prop_fertile))
```

Or the alternative code: 

```
flowers$logitfertile <- car::logit(flowers$prop_fertile)
```

Now let's visualize the transformed data: 

```{r fig.cap = "Normal quantile plot of the proportion of seeds fertilized (logit transformed) on 30 plants", fig.width = 5, fig.height = 4}

flowers %>%
  ggplot(aes(sample = logitfertile)) +
  stat_qq(shape = 1, size = 2) +
  stat_qq_line() +
  ylab("Proportion of seeds fertilized (logit-transformed") +
  xlab("Normal quantile") +
  theme_bw()
```

That's much better!  

Next we learn how to back-transform logit data. 

### Back-transforming logit data {#logit_back}

We'll use the `inv.logit` function from the `boot` package:  

```
?boot::inv.logit
```

First do the back-transform:

```{r logittrans_back}
flowers <- flowers %>%
  mutate(flower_backtransformed = boot::inv.logit(logitfertile))
```

Or alternative code: 

```
flowers$flower_backtransformed <- boot::inv.logit(flowers$logitfertile)
```

Let's have a look at the original "prop_fertile" variable and the "flower_backtransformed" variable to check that they're identical:

```{r shownn}
flowers %>%
  select(prop_fertile, flower_backtransformed)
```

Yup, it worked!

### When to back-transform? {#when_back}

You should back-transform your data when it makes sense to communicate findings on the original measurement scale.  

The most common example is **reporting <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Confidence-interval">confidence intervals</a> for a <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Mean">mean</a> or difference in means**.  

For example, imagine you had calculated a confidence interval for the log-transformed marine biomass ratio data, and your limits were as follows: 

0.347 < $ln(\mu)$ < 0.611

These are the log-transformed limits!  So we need to back-transform them to get them in the original scale:

```{r confback}
lower.limit <- exp(0.347)
upper.limit <- exp(0.611)
```

So now the back-transformed interval is:

`r round(lower.limit, 3)` < $\mu$ < `r round(upper.limit, 3)`

Voila!
