# Creating tables in R Markdown {-}

```{r tablesettings, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")
```

**Tutorial learning objectives**

* Learn how to produce nice tables in R Markdown for PDF output

::: {.callout-note}
**IMPORTANT:  **
This tutorial aims to help you produce nice tables when knitting to PDF.  However, because this tutorial is presented in HTML format, some of the output does not look the same as it will in PDF format.  These instances are noted.
:::

**EXTRA**: there is a new package called [`flextable`](https://ardata-fr.github.io/flextable-book/index.html) that, if you're keen, you are welcome to check out! We don't cover it here.

## Load packages and import data {#tables_packages}

Load the following packages, including `palmerpenguins` and `kableExtra`:

```{r tables_packages, warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)
library(palmerpenguins)
library(kableExtra)
library(biol202)
```

We'll use the `circadian` dataset, which is also used in the ["comparing more than 2 means tutorial"](15-anova.qmd#compare_three_more_means).  These data are associated with Example 15.1 in the text.

The `circadian` data describe melatonin production in 22 people randomly assigned to one of three light treatments.  

```{r anova_getdata}
data(circadian)
```

### Formatting a wide table nicely {#wide_table}

Sometimes a table you want to include in a report is wide — it has many columns — which can make it awkward to read, or a poor fit for a printed page. This often happens with a table of grouped descriptive statistics, of the kind we learned to build in the ["comparing more than 2 means" tutorial](13-two-means.qmd#stats_2samp): each descriptive statistic (count, mean, SD, etc.) becomes its own column, and if you have several of those alongside several groups, the table can end up quite wide. Here we'll learn a general technique for handling this: transposing the table (swapping its rows and columns) and formatting it nicely with the `kbl` function from the `kableExtra` package.

Let's build a table of descriptive statistics for the `circadian` dataset's "shift" variable, grouped by "treatment":

```{r lookdata22}
circadian.stats <- circadian %>%
  group_by(treatment) %>%
  summarise(
    Count = sum(!is.na(shift)),
    Count_NA = sum(is.na(shift)),
    Mean = mean(shift, na.rm = TRUE),
    SD = sd(shift, na.rm = TRUE),
    SEM = SD/sqrt(Count),
    Low_95_CL = t.test(shift, conf.level = 0.95)$conf.int[1],
    Up_95_CL = t.test(shift, conf.level = 0.95)$conf.int[2]
  )
circadian.stats
```

This table has 8 columns (`treatment` plus 7 descriptive statistics) for just 3 treatment groups. Imagine how wide it would get with more groups, or more statistics!

The fix is to "transpose" (rotate) the table using the base `t` function, so that its many columns become rows instead, and its few rows become columns:

```{r transposetable}
circadian.stats.transposed <- t(circadian.stats)
```

Now we'll use the `kbl` function with various arguments to ensure the output looks good and includes a table caption:

::: {.callout-note}
**NOTE:  **
This table will look different when knitted to PDF (rather than HTML)
:::

::: {.callout-note}
**IMPORTANT:  **
Notice that, unlike figure captions, which must be provided in the chunk header, the caption for a table is provided as an argument to the `kbl` function within the actual code!
:::

```
kbl(circadian.stats.transposed, caption = "Descriptive statistics for circadian rhythm shift, by treatment",
    booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

In the preceding code we:

* use the `kbl` function and provide 3 arguments:
  + provide the name of the table-like object we wish to format for output
  + provide a table caption
  + provide an argument "booktabs = TRUE" (this provides nice formatting)

Then we include pipes ("%>%), and follow with:

* the `kable_styling` function with a few arguments:
  + "latex_options = 'hold_position'" which forces the table to appear where the code chunk comes

For lots of great examples of how the `kableExtra` package can be used, see this [vignette](https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf) for PDF output, and [this one](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) for HTML output.

```{r showcircadiantable, echo = FALSE}
kbl(circadian.stats.transposed, caption = "Descriptive statistics for circadian rhythm shift, by treatment",
    booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

Notice one trade-off of transposing: the `treatment`, `Count`, and `Count_NA` rows are whole numbers, but the numeric rows below them (Mean, SD, etc.) now show many decimal places, and there's no way to tell `kbl` to round only some of the rows — transposing converts the whole table to plain text first, so `kbl`'s `digits` argument no longer has any numeric columns to act on. If you need precise control over decimal places, either round the values yourself inside `summarise` before transposing, or use the non-transposed approach in the next section, which keeps that control.

::: {.callout-note}
**TIP: **
If you have many groups and/or many descriptive statistics, the transposed table can still end up wide. In that case, add the "scale_down" option to `kable_styling`, which shrinks the whole table (including the font) to fit the page width:

```
kbl(circadian.stats.transposed, caption = "Descriptive statistics for circadian rhythm shift, by treatment",
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"))
```

For a table with a very large number of columns, this can make the font too small to read comfortably — in that case, consider splitting the table into two, or reporting only the most essential columns.
:::

### A nicely formatted table of descriptive statistics {#stats_nicetable}

Here is the code (which you've already learned) to create a good table of <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Descriptive-statistics">descriptive statistics</a> for a numeric response variable grouped by categories in a categorical explanatory variable.

We'll use the "penguins" dataset again, and calculate descriptive statistics for bill lengths of male penguins, grouped by species:

```{r conf_table_anova}
penguins.stats <- penguins %>%
  filter(sex == "male") %>%
  group_by(species) %>%
  summarise(
    Count = sum(!is.na(bill_length_mm)),
    Count_NA = sum(is.na(bill_length_mm)), 
    Mean = mean(bill_length_mm, na.rm = TRUE),
    SD = sd(bill_length_mm, na.rm = TRUE),
    SEM = SD/sqrt(Count),
    Low_95_CL = t.test(bill_length_mm, conf.level = 0.95)$conf.int[1],
    Up_95_CL = t.test(bill_length_mm, conf.level = 0.95)$conf.int[2]
  )
```

Here's what the raw table looks like:
```{r showsimpletable}
penguins.stats
```

Now let's format it for PDF output:

```{r givenicepdftable}
kbl(penguins.stats, caption = "Descriptive statistics for bill length among male penguins.", 
    booktabs = TRUE, digits = c(0, 0, 0, 2, 3, 3, 3, 3)) %>% 
  kable_styling(latex_options = c("scale_down", "hold_position"), position = "center")
```

The **key difference** from the transposed table in the previous section is that here we specify the number of decimal places we want each descriptive statistic to be reported to, since the table is not transposed and so still has genuine numeric columns for `kbl` to round.

Specifically, the "digits" argument accepts a vector of numbers, whose length is equal to the number of columns being reported in the table, and these numbers indicate the number of decimal places to include for that specific variable.

At this point it would be a good idea to revisit the Biology department's [guidelines for reporting descriptive statistics](https://ubco-biology.github.io/Procedures-and-Guidelines/descriptive-summary-statistics.html). 

For example, we can see that the first three numbers in the "digits" argument are zeroes, and these correspond to the first three columns of the table: "species", "Count", "Count_NA".  These are columns whose values don't require decimal places.

For the "Mean" column we report the values to 1 more decimal place than was used in the measurement (which you find out by looking at the raw data in the "penguins" object), so here, 2 decimal places.

For measures of spread (like the <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Standard-deviation">standard deviation)</a> and measures of uncertainty (including SEM and <a href="https://ubco-biology.github.io/Procedures-and-Guidelines/glossary#Confidence-interval">confidence limits</a>), report the numbers to 2 more decimal places than was used in the measurement, so here, 3 decimal places.

You now know how to produce nicely formatted tables in your knitted PDF output!

::: {.callout-note}
**IMPORTANT:  **
Be sure to try knitting to PDF as soon as you've used any of the `kable` or `kableExtra` package functions in a code chunk, as this will help you trouble-shoot if you encounter problems.
:::
