---
title: "Tutorial work"
author: "Test Student"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

This R Markdown document includes all my notes and coding associated with working through BIOL202 tutorials.

If I want to try out any code during my work, I insert an R "chunk" using the Code > Insert Chunk menu item, or associated shortcut.

Here's an example chunk that tells R to add 2 plus 2. The header of the chunk includes a name "firstchunk"

```{r firstchunk}
2 + 2
```

I will be sure to write issues / tips to myself throughout this document. 

For example: 

**REMEMBER** You only need to install a package once!  You do need to "load" a package each time you start R/RStudio.

I can consult the R Markdown [reference guide ](https://utstat.utoronto.ca/reid/sta2201s/rmarkdown-reference.pdf) for info about how to format my R Markdown document.

And if I ever wish to see what all this work looks like in HTML (webpage) format, I could simply select the "knit" button at the top and knit to HTML (or PDF if I wish).  But knitting is more important for my markdown document that i'll have for assignments!

I will provide a link to the relevant tutorial for each section.  Sections are denoted with headers using two hashtags.

Here we go!

## Getting started with R and RStudio

The tutorials in this section are linked [here](https://ubco-biology.github.io/BIOL202-lab-tutorials/01-getting-started.html).

Boy, this was challenging to get started! I got hung up on ....



